My wonderful paper

### 2024-11-07 _manuscript/_tex/index.tex
remove the result universe example

(might be useful) Analysis plans can be constructed at various
granularities, at the highest level, one may only has a plan of the
specific method used for analysing data and the expected outcome. This
provides little guidance when a deviation from expectation occurs. At
the lowest level, one may have a plan for each data entry and every data
handling steps. This provides too much detail and may not be practical
in practice.

@@

Furthermore, by generating multiple versions of the data, we can emulate
various scenarios within the same context for students to exercise
judgement when conducting data analysis in a classroom setting.

@@

To build the result universe, datasets can be simulated to either meet
and fail these plan expectations, allowing the analysts to observe the
significance of PM10 coefficient. Here, sample sizes of 50, 100, 500,
and 1000 are considered. Two model specifications are included: 1)
linear model with PM10 as the only covariate
(\(\text{mortality} \sim \text{PM10}\)), 2) linear model with PM10 and
temperature as covariates
(\(\text{mortality} \sim \text{PM10} + \text{temp}\)). A grid-based
approach is used to simulate correlation structure. Reasonable ranges of
correlation between the three variables are
\(\text{cor}(\text{mortality}, \text{PM10}) \in [-0.01, 0]\),
\(\text{cor}(\text{mortality}, \text{temperature}) \in [-0.6, -0.2]\),
and \(\text{cor}(\text{PM10}, \text{temperature}) \in [0.2, 0.6]\).

\begin{itemize}
\item
  add a paragraph to describe the simulation process
\item
  add a fourth panel to describe the comparison of a right/ wrong
  expectation, i.e.~correlation on PM10 and mortality
\end{itemize}

Figure~\ref{fig-result-universe} shows that result universe of the
linear regression model and how a change of decision in one of the plan
expectations above affect the outcome expectation. Panel a) is colored
by the outcome expectation -- whether a significant p-value is found in
the PM10 coefficient. Panel b) shows the effect of adding temperature to
the model and the results show that the significance of PM10 coefficient
can be achieved by adding temperature to the model for a sample size of
500. Panel c) shows that increasing sample size from 50 to 100 enhances
the significance of p-value for PM10 and the significance remains with
further increases in sample size. {[}note: weave the ``actual data''
into the example linear regression model{]}

\phantomsection\label{cell-fig-result-universe}
\begin{figure}[H]

\centering{

\includegraphics{index_files/figure-pdf/fig-result-universe-1.pdf}

}

\caption{\label{fig-result-universe}The result universe of linear
regression model to study the effect of PM10 on mortality: a) colored by
whether the p-value of PM10is significant (less than 0.05), b) the
effect of adding temperature to the model for a sample size of 500, c)
the effect of increasing sample size for a fixed correlation structure.}

\end{figure}%

@@

\begin{itemize}
\item
  how to systematically simulate data is still unknown, sensitivity of
  the simulation to the results
\item
  plotting is a critical way to check data and they can still be frame
  into a unit test. it is a open problem to how to encode the
  visualization into the unit tests. Maybe a procedure like confirm plot
  (this looks alright to you) and then press the button to continue
\item
  currently no automated way to generate unit tests. It is interesting
  to see the automation of generating unit tests, although it requires
  the inputs from experts across a wide array of common scenarios.
\item
  There are cost and benefit on what granularity to set expectation.
  Setting them on every step of the analysis requires more work from the
  analysts while for complex analysis, the diagnostics may require
  divide the analysis into different components and diagnose it in a
  hierarchical structure. The method proposed in this paper should be
  applicable for both fine and coarse analysis.
\item
  (only mention it) Software testing relies on ``oracles'' to provide
  the expected output necessary for verifying test results. For example,
  to test whether the program correctly calculates 1 + 1, one need to
  supply the correct answer, 2. However, establishing this ``correct''
  output can sometimes be challenging, where obtaining a solution may be
  difficult without the program itself. This situation leads to the
  oracle problem \citep{barr2014oracle}. In data analysis, the similar
  oracle problem can happen, as the ``truth'' of an outcome, the
  expectation, depends on the underlying theory or interpretation. For
  example, in a linear regression model, the significance of a
  coefficient may be expected or unexpected based on the theory, making
  it challenging for researcher with a different theory to assess the
  results and the analysis.
\end{itemize}

>>>>

@@

@@

\begin{itemize}
\tightlist
\item
  add a paragraph to describe the simulation process
\end{itemize}

@@

\begin{itemize}
\item
  how to systematically simulate data is still unknown, sensitivity of
  the simulation to the results
\item
  plotting is a critical way to check data and they can still be frame
  into a unit test. it is a open problem to how to encode the
  visualization into the unit tests. Maybe a procedure like confirm plot
  (this looks alright to you) and then press the button to continue
\item
  currently no automated way to generate unit tests. It is interesting
  to see the automation of generating unit tests, although it requires
  the inputs from experts across a wide array of common scenarios.
\item
  There are cost and benefit on setting expectation on different
  granularity. At the lowest level, one may have a plan for each data
  entry and every data handling steps. This requires more work from the
  analysts and may not be practical in practice. For more complex
  analyses, analysts may divide the analysis into sections and set
  expectations for each. They can then focus on the specific sections
  flagged by the tests and sub-divide the sections to set expectation
  and diagnose the analysis in a hierarchical manner.
\item
  (only mention it) Software testing relies on ``oracles'' to provide
  the expected output necessary for verifying test results. For example,
  to test whether the program correctly calculates 1 + 1, one need to
  supply the correct answer, 2. However, establishing this ``correct''
  output can sometimes be challenging, where obtaining a solution may be
  difficult without the program itself. This situation leads to the
  oracle problem \citep{barr2014oracle}. In data analysis, the similar
  oracle problem can happen, as the ``truth'' of an outcome, the
  expectation, depends on the underlying theory or interpretation. For
  example, in a linear regression model, the significance of a
  coefficient may be expected or unexpected based on the theory, making
  it challenging for researcher with a different theory to assess the
  results and the analysis.
\end{itemize}

### 2024-11-07 index.qmd
remove the result universe example

To build the result universe, datasets can be simulated to either meet and fail these plan expectations, allowing the analysts to observe the significance of PM10 coefficient. Here, sample sizes of 50, 100, 500, and 1000 are considered. Two model specifications are included: 1) linear model with PM10 as the only covariate ($\text{mortality} \sim \text{PM10}$), 2) linear model with PM10 and temperature as covariates ($\text{mortality} \sim \text{PM10} + \text{temp}$). A grid-based approach is used to simulate correlation structure. Reasonable ranges of correlation between the three variables are $\text{cor}(\text{mortality}, \text{PM10}) \in [-0.01, 0]$, $\text{cor}(\text{mortality}, \text{temperature}) \in [-0.6, -0.2]$, and $\text{cor}(\text{PM10}, \text{temperature}) \in [0.2, 0.6]$.

@@

 * add a fourth panel to describe the comparison of a right/ wrong expectation, i.e. correlation on PM10 and mortality

@fig-result-universe shows that result universe of the linear regression model and how a change of decision in one of the plan expectations above affect the outcome expectation. Panel a) is colored by the outcome expectation -- whether a significant p-value is found in the PM10 coefficient. Panel b) shows the effect of adding temperature to the model and the results show that the significance of PM10 coefficient can be achieved by adding temperature to the model for a sample size of 500. Panel c) shows that increasing sample size from 50 to 100 enhances the significance of p-value for PM10 and the significance remains with further increases in sample size. [note: weave the "actual data" into the example linear regression model]

@@

# test for dist fit
# fitdist(pm10$pm10, "gamma") -> pm10_fit
# plot(pm10_fit) # gamma is better than normal for the fit of pm10
#fitdist(pm10$temp, "norm") -> temp_fit
# plot(temp_fit)

@@

```{r}
#| label: fig-result-universe
#| fig-cap: "The result universe of linear regression model to study the effect of PM10 on mortality: a) colored by whether the p-value of PM10is significant (less than 0.05), b) the effect of adding temperature to the model for a sample size of 500, c) the effect of increasing sample size for a fixed correlation structure."
#| fig.width: 12
#| fig.height: 12
dt <- res |>   
  mutate(     
    p_value = fit[2,4],     
    coef = fit[2,1],     
    xy_correlation = as.numeric(corr_mtx[1, 2]),     
    xz_correlation = as.numeric(corr_mtx[1, 3]),     
    yz_correlation = as.numeric(corr_mtx[2, 3]),     
    fml = deparse(model) |> as.factor(),     
    expect = ifelse(p_value < 0.05, 1, 0) |> as.factor(),     
    ) |>    
  ungroup()

code_tbl <- crossing(V1 = c("signifpm10", NA),           
                     V2 = c("smallsample", NA), 
                     V3 = c("withtemp", NA)) |>      
  mutate(col1 = paste0(V1,"_", V2, "_", V3))

lookup_tbl <- tibble(   
  value = crossing(x = c(1, 0), y = c(1, 0), z = c(1, 0)) |>      
    mutate(col = paste0(x,"_", y, "_", z)) |> 
    arrange(-x, -y, -z) |>      
    pull(col),   
  plan1 = unique(code_tbl$col1) )

dt2 <- dt |>    
  mutate(T1 = ifelse(sample_size < 200, 1, 0),          
         T2 = ifelse(fml == "mortality ~ pm10 + temp", 1, 0),          
         pp1 = paste0(expect, "_", T1, "_", T2),) |>    
  left_join(lookup_tbl, by = c("pp1" = "value"))

p0 <- dt |> ggplot() +   
  geom_point(aes(x = p_value , y = coef), color = "grey", size = 0.5) +  
  labs(x = "p-value", y = "coefficient") +   
  theme_bw() +    
  theme(aspect.ratio = 1, 
        panel.grid.minor = element_blank())

df0 <-  dt |> filter(sample_size == 500, xy_correlation == -0.005, 
                     between(xz_correlation, -0.46, -0.43),
                     yz_correlation == 0.4) |> arrange(expect)
df2 <- df0 |> filter(fml == "mortality ~ pm10 + temp") # test for model spec

p1 <- p0 + 
  geom_point(aes(x = p_value , y = coef, color = expect), size = 0.3) +  
  scale_color_brewer(palette = "Dark2", direction = -1) +
  labs(title = "Coefficient vs. p-value") + 
  theme(legend.position = 'none')

p2 <- p0 + 
  geom_point(data = df0, aes(x = p_value, y = coef, color = expect), size = 2) +
  geom_path(data = df0, aes(x = p_value, y = coef), color = "black",
            arrow = arrow(type = "open", angle = 30, 
                          length = unit(0.1, "inches"))) +
  ggrepel::geom_label_repel(
    data = df0, aes(x = p_value, y = coef, label = fml, color = expect)
    ) +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  labs(title = "Effect of adding temperature")

df00 <-  dt |> filter(xy_correlation == -0.005, 
                     between(xz_correlation, -0.46, -0.43),
                     yz_correlation == 0.4) |> arrange(expect)
df <- df00 |> filter((fml == "mortality ~ pm10 + temp"))

p3 <- p0 + 
  geom_point(data = df, aes(x = p_value, y = coef, color = expect), size = 2) +
  geom_path(data = df, aes(x = p_value, y = coef), color = "black",
            arrow = arrow(type = "open", angle = 30, 
                          length = unit(0.1, "inches"))) +
  ggrepel::geom_label_repel(
    data = df, aes(x = p_value, y = coef, label = sample_size, color = expect)
    ) +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  labs(title = "Effect of sample size")

(p1 | p2) / (p3 | plot_spacer()) + 
  plot_annotation(tag_levels = "a") + 
  plot_layout(guides = 'collect') &
  theme(legend.position = 'bottom') 
```

@@

* There are cost and benefit on what granularity to set expectation. Setting them on every step of the analysis requires more work from the analysts while for complex analysis, the diagnostics may require divide the analysis into different components and diagnose it in a hierarchical structure. The method proposed in this paper should be applicable for both fine and coarse analysis.

>>>>

@@

@@

@@

@@

* There are cost and benefit on setting expectation on different granularity. At the lowest level, one may have a plan for each data entry and every data handling steps. This requires more work from the analysts and may not be practical in practice. For more complex analyses, analysts may divide the analysis into sections and set expectations for each. They can then focus on the specific sections flagged by the tests and sub-divide the sections to set expectation and diagnose the analysis in a hierarchical manner.
